---
title: "Reproducible Research Course Project 1"
output: html_document
---

The data for the assignment can be downloaded [here] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). 

## Loading and preprocessing the data

```{r echo = TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(plyr)
activity <- read.csv("activity.csv")
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime <- as.POSIXct(activity$date, format="%Y-%m-%d")
```

pulling data without nas

```{r echo = TRUE}
clean <- activity[!is.na(activity$steps),]
```

## What is mean total number of steps taken per day?

summarizing total steps per date

```{r echo = TRUE}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```

Creating the historgram of total steps per day

```{r histogram1, echo=TRUE, fig.width=10, warning=FALSE}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
as.integer(mean(sumTable$Steps))
as.integer(median(sumTable$Steps))
```

## What is the average daily activity pattern?

pulling data without nas

``{r echo = TRUE)
clean <- activity[!is.na(activity$steps),]
```
create average number of steps per interval

```{r echo = TRUE}
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))
```

Create line plot of average number of steps per interval

```{r timeplot1, echo=TRUE, fig.width=10, warning=FALSE}
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", 
            ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+
        ggtitle("Average Number of Steps per Interval")
```

Mean and median of total number of steps taken per day

```{r echo = TRUE}
maxSteps <- max(intervalTable$Avg)
intervalTable[intervalTable$Avg==maxSteps,1]
```

## Imputing missing values

Number of NAs in original data set

```{r echo = TRUE}
nrow(activity[is.na(activity$steps),])
```

Create the average number of steps per weekday and interval

```{r echo = TRUE}
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
nadata<- activity[is.na(activity$steps),]
```

Merge NA data with average weekday interval for substitution

```{r echo = TRUE}
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```

Reorder the new substituded data in the same format as clean data set

```{r echo = TRUE}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
```

Merge the NA averages and non NA data together

```{r echo = TRUE}
mergeData <- rbind(clean, newdata2)
```

Create sum of steps per date to compare with step 1

```{r echo = TRUE}
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")
```

Mean and Median of Steps with NA data taken care of

```{r echo = TRUE}
as.integer(mean(sumTable2$Steps))
as.integer(median(sumTable2$Steps))
```

Creating the histogram of total steps per day, categorized by data set to show impact

```{r histogram2, echo=TRUE, fig.width=10, warning=FALSE}
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", 
     col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```

## Are there differences in activity patterns between weekdays and weekends?

Create new category based on the days of the week

```{r echo = TRUE}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
library(lattice) 
```

Summarize data by interval and type of day

```{r echo = TRUE}
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))
```

Plot data in a panel plot

```{r timeplot2, echo=TRUE, fig.width=10, warning=FALSE}
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```